---
title: "Ferritin meta-regression"
author: "John J Davis, PhD"
date: "2024-03-13"
output: html_document
---

```{r setup, include=FALSE, echo=FALSE, warning=FALSE}


knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(metafor)
library(cowplot)
library(knitr)
library(mgcv)


```

## Meta-regression of ferritin data from Smid et al. 2024


[Smid et al. 2024](https://link.springer.com/article/10.1007/s40279-024-01992-8) provide data from 13 studies on oral iron supplementation in athletes, totaling N=449 subjects.  Data from this meta-analysis were aggragated using tables in the intial publication and digitization of scatterpoints in Figure 8 in the manuscript.  


```{r read_data}

#Read data and display a glimpse
df <- read_csv("smid_data.csv", show_col_types = FALSE)
df %>% glimpse()


```

## Intial ferritin and ferritin increase from iron supplements

First, reproduce the unstratified meta-analysis of Smid et al. to ensure we get the same results. 


```{r ferritin_forest}

#Smid et al. used SPSS, whose default method is REML random effects
meta <- rma(ferritin_effect_size_smd, sei = ferritin_std_error_smd, slab=id, 
            data = df, method="REML", measure="SMD") 
meta


# --- Make forest plot ----
forest(meta, header="Study", mlab="Overall pooled results", cex=1.0,
       main="Effect of iron supplements on ferritin level",
       cex.main = 1.5)


```

Smid et al. report an overall effect size of 1.27 (0.44, 2.10); the minor differences in the 95% confidence intervals of this estimate are likely due to implementation differences in REML estimation in `metafor` vs. SPSS.  


### Semi-parametric meta-regression

Visual inspection of the effect size vs. initial ferritin level suggets a smooth, nonlinear relationship:  

```{r ferritin_bubble, echo=FALSE}

#Plot parameters
point_fill <- "#deebf7"
point_color <- "black"

lwd <- 1
draw_lwd <- 0.5
fnt <- 14
title_fnt <- 14
subtitle_fnt <- 12
wm_fnt <- 12
hline_lwd <- 0.75
grid_lwd <- 0.5
alf <- 0.5
bubble_min <- 2
bubble_max <- 9


plt <- df %>%
  #Scale bubble by inverse of standard error
  ggplot(aes(x=initial_ferritin_ng_ml, y=ferritin_effect_size_smd, size=2/ferritin_std_error_smd)) +
  #Study bubbles
  geom_point(shape=21,
             fill = point_fill,
             stroke = 0.2,
             color = point_color) + 
  #Change point sizing
  scale_size(guide="none", range=c(bubble_min,bubble_max)) +
  #Set axis limits
  scale_x_continuous(limits = c(0,50), expand = c(0,0),
                     breaks = seq(0,50, by=5),
                     name = "Ferritin level (ng/mL)") + 
  scale_y_continuous(limits = c(-1,8), expand = c(0,0),
                     breaks = seq(-2,8,by=1),
                     name = "Ferritin increase (effect size)") + 
  coord_cartesian(ylim = c(-1,6.5))+
  ggtitle(label = "Ferritin increase from iron supplements, as a function of initial ferritin level", 
          subtitle = "Data from Smid et al. 2024") + 
  theme_bw() + 
  theme(plot.title = element_text(hjust = 0.5, size=title_fnt),
        plot.caption.position = "panel",
        plot.subtitle = element_text(hjust = 0.5, size=subtitle_fnt),
        panel.grid.major = element_line(linewidth = grid_lwd),
        panel.grid.minor = element_blank(),
        axis.title = element_text(size = fnt, color = "black"),
        axis.text = element_text(size = fnt, color = "black"))
plt



```


The code below implements a moderator analysis, which treats inital serum ferritin as a smooth and potentially nonlinear moderator of the effects of iron supplementation on serum ferritin increases. The effect is modeled using a cubic regression spline basis, as implemented in `mgcv` and the meta-regression package `metafor`. The analysis below follows the [official metafor tutorial documentation here.](https://www.metafor-project.org/doku.php/tips:non_linear_meta_regression) See the end of this document for a sensitivity analysis of `K`, the number of knots in the spline basis.  


```{r ferritin_metareg}

#X-grid for plotting
xs <- seq(5, 50, length=500)

# --- Linear fit, purely for plotting purposes --- 
# do note we pass sei (standard errors) here, not the default variance
# See https://www.metafor-project.org/doku.php/tips:input_to_rma_function for details
res.lin <- rma(ferritin_effect_size_smd, sei = ferritin_std_error_smd, slab=id, 
               mods = ~ initial_ferritin_ng_ml,
               data = df, method="REML", measure="SMD")

# --- Smooth nonlinear meta-regression ---
# Adapted from https://www.metafor-project.org/doku.php/tips:non_linear_meta_regression

#Number of knots for spline fit, using sqrt(N) approximate rule (Harezlak et al.)
K <- 4 #See later for sensitivity analysis


#Construct a cubic spline basis in mgcv with K knots
sm <- smoothCon(s(initial_ferritin_ng_ml, bs="cr", k=K), data=df, absorb.cons=TRUE)[[1]]

#Do REML-based meta-regression, using the spline basis above as moderator
res_fer <- rma(ferritin_effect_size_smd, sei = ferritin_std_error_smd, slab=id, 
               mods = ~ sm$X,
               data = df, method="REML", measure="SMD")
#Show results
res_fer

#Make raw plot using metafor's base R plotting capabilities
sav <- predict(res_fer, newmods=PredictMat(sm, data.frame(initial_ferritin_ng_ml=xs)))
regplot(res.lin, mod=2, pred=sav, xvals=xs, las=1, digits=1, bty="l",
        psize=.75/df$ferritin_std_error_smd, 
        xlab="Initial ferritin (ng/mL)", main="Cubic spline model")


```


Note that because of the weighting, smaller studies have less of an influence on the spline fit.  

The results of the moderator analysis strongly suggest that initial ferritin level is a significant moderator of the effects of iron supplementation on serum ferritin increase (omnibus test p-value: `r sprintf("%1.2e", res_fer$QMp)`).  



## Initial ferritin level and VO2max

We can also conduct a similar analysis for VO2max gains after iron supplementation. Note that only ten of the 13 studies reported VO2max data, hence the `NA` warning below.  


Again, we'll start by replicating the meta-analysis and forest plots.  


```{r vo2max_forest}


meta_vo2 <- rma(vo2max_effect_size, sei = vo2max_std_error, slab=id, 
            data = df, method="REML", measure="SMD")
meta_vo2

# --- Make forest plot ----
forest(meta_vo2, header="Study", mlab="Overall pooled results", cex=1.0,
       main="Effect of iron supplements on VO2max",
       cex.main = 1.5)



```

Smid et al. report an overall pooled result of 0.49 (-0.08, 1.07).  


### Semi-parametric meta-regression

Again, a visual inspection of the VO2max effect size vs. initial ferritin level suggets a smooth, nonlinear relationship:  




```{r vo2max_bubble, echo=FALSE, warning=FALSE}


#Plot parameters
point_fill <- "#bcbddc"
point_color <- "black"


plt <- df %>%
  #Scale bubble by inverse of standard error
  ggplot(aes(x=initial_ferritin_ng_ml, y=vo2max_effect_size, size=2/vo2max_std_error)) +
  #Study bubbles
  geom_point(shape=21,
             fill = point_fill,
             stroke = 0.2,
             color = point_color) + 
  #Change point sizing
  scale_size(guide="none", range=c(bubble_min,bubble_max)) +
  #Set axis limits
  scale_x_continuous(limits = c(0,50), expand = c(0,0),
                     breaks = seq(0,50, by=5),
                     name = "Ferritin level (ng/mL)") + 
  scale_y_continuous(limits = c(-1,8), expand = c(0,0),
                     breaks = seq(-2,8,by=1),
                     name = "VO2max increase (effect size)") + 
  coord_cartesian(ylim = c(-1,6.5))+
  ggtitle(label = "VO2max increase from iron supplements, as a function of initial ferritin level", 
          subtitle = "Data from Smid et al. 2024") + 
  theme_bw() + 
  theme(plot.title = element_text(hjust = 0.5, size=title_fnt),
        plot.caption.position = "panel",
        plot.subtitle = element_text(hjust = 0.5, size=subtitle_fnt),
        panel.grid.major = element_line(linewidth = grid_lwd),
        panel.grid.minor = element_blank(),
        axis.title = element_text(size = fnt, color = "black"),
        axis.text = element_text(size = fnt, color = "black"))
plt



```

The code below implements a similar semiparametric moderator analysis as above.  


```{r vo2max_metareg}

#Don't use all data for spline basis, just studies with VO2 measurements
df_vo2 <- df %>%
  drop_na(vo2max_effect_size)


#Plotting setup 
xs_vo2 <- seq(5, 35, length=500) #no VO2 data for the higher ranges

# --- Linear fit, purely for plotting purposes --- 
res.lin_vo2 <- rma(vo2max_effect_size, sei = vo2max_std_error, slab=id, 
               mods = ~ initial_ferritin_ng_ml,
               data = df_vo2, method="REML", measure="SMD")

# --- Smooth nonlinear meta-regression ---

K <- 3 #Fewer studies, fewer knots
sm <- smoothCon(s(initial_ferritin_ng_ml, bs="cr", k=K), data=df_vo2, absorb.cons=TRUE)[[1]]
res_vo2 <- rma(vo2max_effect_size, sei = vo2max_std_error, slab=id, 
               mods = ~ sm$X,
               data = df_vo2, method="REML", measure="SMD")
res_vo2

sav_vo2 <- predict(res_vo2, newmods=PredictMat(sm, data.frame(initial_ferritin_ng_ml=xs_vo2)))
regplot(res.lin_vo2, mod=2, pred=sav_vo2, xvals=xs_vo2, las=1, digits=1, bty="l",
        psize=.75/df_vo2$vo2max_std_error, 
        xlab="Initial ferritin (ng/mL)", main="Cubic spline model")


```


As above, the results of the moderator analysis strongly suggest that initial ferritin level is a significant moderator of the effects of iron supplementation on VO2max increase (omnibus test p-value: `r sprintf("%.4f", res_vo2$QMp)`).  

Notably, the borderline significant results from the standard pooled analysis earlier (p = `r sprintf("%.4f", meta_vo2$pval)`) are now definitively significant at the 0.05 level.   



```{r publication_plots, echo=FALSE}


#Get p-values for omnibus test of moderator effect
# (null hypothesis: moderator curve is zero everywhere)
ferritin_omni_p <- res_fer$QMp
vo2_omni_p <- res_vo2$QMp

ferritin_ant_p <- "p < 0.001" #10e-13...
vo2_ant_p <- sprintf("p = %.3f", vo2_omni_p)


#Fetch meta-regression results for publication plots
met_plot <- data.frame(ferritin = xs,
                      yhat = sav$pred,
                      ci_lo = sav$ci.lb,
                      ci_hi = sav$ci.ub)

met_plot_vo2 <- data.frame(ferritin = xs_vo2,
                       yhat = sav_vo2$pred,
                       ci_lo = sav_vo2$ci.lb,
                       ci_hi = sav_vo2$ci.ub)


#Combine and make publication-ready plots

#6.5" by ahfl that? 3.25


# --- Ferritin plot ---



point_fill <- "#deebf7"
point_color <- "black"

lwd <- 0.75
draw_lwd <- 0.15
fnt <- 8
title_fnt <- 8
subtitle_fnt <- 6
wm_fnt <- 7
hline_lwd <- 0.25
grid_lwd <- 0.2

alf <- 0.2


bubble_min <- 0.75
bubble_max <- 4

ant_fnt <- 3
ant_x <- 49.5
ant_y <- 6.4

#, range=c(0.5,4)

plt_1 <- met_plot %>%
  ggplot(aes(x=ferritin, y=yhat)) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = hline_lwd) + 
  geom_ribbon(aes(ymin = ci_lo, ymax = ci_hi), fill = "#3182bd", alpha = alf) + 
  geom_line(color = "#3182bd", linewidth = lwd) + 
  #Omnibus moderator test p-value
  annotate("text", x=ant_x, y=ant_y, label = ferritin_ant_p, size=ant_fnt,
           hjust = 1, vjust=1) + 
  #Study bubbles
  geom_point(aes(x=initial_ferritin_ng_ml , y=ferritin_effect_size_smd, size=2/ferritin_std_error_smd ),
             data = df, 
             shape=21,
             fill = point_fill,
             stroke = 0.2,
             color = point_color) + 
  
  #Change point sizing
  scale_size(guide="none", range=c(bubble_min,bubble_max)) +
  #Set axis limits
  scale_x_continuous(limits = c(0,50), expand = c(0,0),
                     breaks = seq(0,50, by=5),
                     name = "Initial serum ferritin level (\u03bcg/L)") +  
  scale_y_continuous(limits = c(-1,8), expand = c(0,0),
                     breaks = seq(-2,8,by=1),
                     name = "Serum ferritin increase (SMD)") + 
  coord_cartesian(ylim = c(-1,6.5))+
  ggtitle(label = "Serum ferritin increase from oral iron supplementation") +
  theme_bw() + 
  theme(plot.title = element_text(hjust = 0.5, size=title_fnt),
        plot.caption.position = "panel",
        plot.caption = element_text(face = "bold", size = wm_fnt, vjust=0),
        plot.subtitle = element_text(hjust = 0.5, size=subtitle_fnt),
        panel.grid.major = element_line(linewidth = grid_lwd),
        panel.grid.minor = element_blank(),
        axis.title = element_text(size = fnt, color = "black"),
        axis.text = element_text(size = fnt, color = "black"))

plt_1


# --- VO2max ---
point_fill <- "#efedf5"
point_color <- "black"

line_col <- "#756bb1"
rib_col <- "#756bb1"


plt_2 <- met_plot_vo2 %>%
  ggplot(aes(x=ferritin, y=yhat)) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = hline_lwd) + 
  geom_ribbon(aes(ymin = ci_lo, ymax = ci_hi), fill = rib_col, alpha = alf) + 
  geom_line(color = line_col, linewidth = lwd) + 
  #Omnibus moderator test p-value
  annotate("text", x=ant_x, y=ant_y, label = vo2_ant_p, size=ant_fnt,
           hjust = 1, vjust=1) + 
  #Study bubbles
  geom_point(aes(x=initial_ferritin_ng_ml , y=vo2max_effect_size, size=2/vo2max_std_error),
             data = df_vo2, 
             shape=21,
             fill = point_fill,
             stroke = 0.2,
             color = point_color) + 
  #Change point sizing
  scale_size(guide="none", range=c(bubble_min,bubble_max)) +
  #Set axis limits
  scale_x_continuous(limits = c(0,50), expand = c(0,0),
                     breaks = seq(0,50, by=5),
                     name = "Initial serum ferritin level (\u03bcg/L)") +  
  scale_y_continuous(limits = c(-2,8), expand = c(0,0),
                     breaks = seq(-2,8,by=1),
                     name = "VO2max increase (SMD)") + 
  coord_cartesian(ylim = c(-1,6.5))+
  ggtitle(label = "VO2max increase from oral iron supplementation") +
  theme_bw() + 
  theme(plot.title = element_text(hjust = 0.5, size=title_fnt),
        plot.caption.position = "panel",
        plot.caption = element_text(face = "bold", size = wm_fnt, vjust=0),
        plot.subtitle = element_text(hjust = 0.5, size=subtitle_fnt),
        panel.grid.major = element_line(linewidth = grid_lwd),
        panel.grid.minor = element_blank(),
        axis.title = element_text(size = fnt, color = "black"),
        axis.text = element_text(size = fnt, color = "black"))

plt_2


#Combine and save
cow <- plot_grid(plt_1, plt_2, nrow=1)
ggsave("Figure 1 - Moderating effects of initial ferritin on effects of iron supplementation in athletes.png", 
       plot=cow, width = 6.5, height = 3, units = "in", dpi=600)



```



## Knot number sensitivity analysis


Assess whether choice of `K` has a strong bearing on the shape of the results.  

```{r sens_analysis}

#Eval grid
xs <- seq(5, 50, length=500)
K_vals <- c(3,4,5,6) #Knots to test

#Define function to use in loop
fit_mod <- function(yi, sei, mod_vals, K, xs){
  #Do semipar meta-regression, evaluate on grid, and return evaluated curve
  df_dummy <- data.frame(yi = yi, 
                         sei = sei,
                         mod_vals = mod_vals)
  #Spline basis
  sm <- smoothCon(s(mod_vals, bs="cr", k=K), data=df_dummy, absorb.cons=TRUE)[[1]]
  
  res_spline <- rma(yi=yi, sei=sei,
                    mods = ~ sm$X,
                    data = df_dummy, method="REML", measure="SMD")
  sav_spline <- predict(res_spline, newmods=PredictMat(sm, data.frame(mod_vals=xs)))
  
  plot_df <- data.frame(mod_vals = xs,
                      yhat = sav_spline$pred,
                      ci_lo = sav_spline$ci.lb,
                      ci_hi = sav_spline$ci.ub,
                      K=K)
  return(plot_df)
}

# --- Ferritin increase sensitivity ---
mod_vals <- df$initial_ferritin_ng_ml
yi <- df$ferritin_effect_size_smd
sei <- df$ferritin_std_error_smd

result_list <- list()

for (i in 1:length(K_vals)){
  result_list[[i]] <- fit_mod(yi, sei, mod_vals, K_vals[i], xs)
}

ferritin_sens <- bind_rows(result_list) %>% mutate(K = factor(K))

ferritin_sens_plot <- ferritin_sens %>%
  ggplot(aes(x=mod_vals, y=yhat, color=K)) + 
  geom_line(linewidth=1.5, alpha = 0.7) + 
  scale_x_continuous(limits = c(0,50)) + 
  ggtitle("K sensitivity analysis - ferritin increase")

ferritin_sens_plot


# --- VO2max ---


#Eval grid
xs <- seq(5, 35, length=500)
K_vals <- c(3,4,5) #10 studies, no sense using 6 knots

#Data
mod_vals <- df_vo2$initial_ferritin_ng_ml
yi <- df_vo2$vo2max_effect_size
sei <- df_vo2$vo2max_std_error

result_list <- list()

for (i in 1:length(K_vals)){
  result_list[[i]] <- fit_mod(yi, sei, mod_vals, K_vals[i], xs)
}

vo2_sens <- bind_rows(result_list) %>% mutate(K = factor(K))

vo2_sens_plot <- vo2_sens %>%
  ggplot(aes(x=mod_vals, y=yhat, color=K)) + 
  geom_line(linewidth=1.5, alpha = 0.7) +  
  scale_x_continuous(limits = c(0,50)) + 
  ggtitle("K sensitivity analysis - VO2max increase")

vo2_sens_plot


```


For increased serum ferritin, the choice of K does not have a particularly strong effect on the results in the 10-25 ng/mL range. `K=3` is likely too small, given its less-plausible U-shaped curve.  

For increased VO2max, the choice of K has a modest effect on estimated benefits of iron supplementation in the 10-25 ng/mL range. All three choices still show a definite increase in VO2max at low levels of ferritin, with the "wiggles" near 15 ng/mL for `K=4` and `K=5` likely attributable to the dearth of studies with initial ferrtin levels of 17-25 ng/mL.  



